---
title: "Hands-on Exercise 3: Spatial Point Patterns Analysis"
date: "29 january 2023"
date-modified: "`r Sys.Date()`"
number-sections: true
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

## Introduction

Spatial Point Pattern Analysis is the evaluation of the pattern or distribution, of a set of points on a surface.

## Installing and Loading R packages

```{r}
pacman::p_load(maptools, sf, raster, spatstat, tmap)
```

## Spatial Data Wrangling

We first use st_red() to import geospatial data into R

```{r}
childcare_sf <- st_read("data/child-care-services-geojson.geojson") %>%
  st_transform(crs = 3414)
```

```{r}
sg_sf <- st_read(dsn = "data", layer="CostalOutline")
```

```{r}
mpsz_sf <- st_read(dsn = "data", 
                layer = "MP14_SUBZONE_WEB_PL")
```

Using the appropriate **sf** function you learned in Hands-on Exercise 2, retrieve the referencing system information of these geospatial data.

```{r}
st_crs(childcare_sf)

st_crs(sg_sf)

st_crs(mpsz_sf)
```

Notice that except childcare_sf, both mpsz_sf and sg_sf do not have proper crs information.

Using the method you learned in Lesson 2, assign the correct crs to mpsz_sf and sg_sf simple feature data frames.

```{r}
sg_sf<-st_transform(sg_sf, crs = 3414)

mpsz_sf<-st_transform(mpsz_sf, crs = 3414)
```

DIY: create map as given

```{r}
tmap_mode("plot")

qtm(mpsz_sf)+
qtm(childcare_sf)
```

Alternatively, we can prepare a pin map by using the code chunk below. tmap mode is set to view: interactive pin map: allows us to navigate and zoom around the map freely.

```{r}
tmap_mode('view')
tm_shape(childcare_sf)+
  tm_dots()
```

Set map mode back to plot, to avoid having an excessive number of interactive maps in one document

```{r}
tmap_mode('plot')
```

## Geospatial Data Wrangling

Converting sf data frames to sp's spatial class

```{r}
childcare <- as_Spatial(childcare_sf)
mpsz <- as_Spatial(mpsz_sf)
sg <- as_Spatial(sg_sf)
```

DIY: Using appropriate function, display the information of these three Spatial\* classes as shown below.

```{r}
childcare
mpsz
sg
```

## Converting the Spatial class unto generic sp format

spatstat requires the analytical data in ppp object form

```{r}
childcare_sp <- as(childcare, "SpatialPoints")
sg_sp <- as(sg, "SpatialPolygons")
```

```{r}
childcare_sp
sg_sp
```

### Converting the genetic sp format into spatstat's ppp format

```{r}
childcare_ppp <- as(childcare_sp, "ppp")
childcare_ppp
```

plotting childcare_ppp

```{r}
plot(childcare_ppp)
```

Taking a quick look at summary statistics of newly created ppp object

```{r}
summary(childcare_ppp)
```

Notice the warning message about duplicates. In ppp analysis, an issue of significance is the presence of duplicates. Points cannot be coincident.

### Handling duplicated points

Checking for any duplicated objects

```{r}
any(duplicated(childcare_ppp))
```

To count the number of co-indicence points, we use the mutiplicity() function

```{r}
multiplicity(childcare_ppp)
```

Counting the number of duplicated points

```{r}
sum(multiplicity(childcare_ppp) > 1)
```

To view the locations of dupicated point events, plot childcare data by using code below

```{r}
tmap_mode('view')
tm_shape(childcare) +
  tm_dots(alpha=0.4, 
          size=0.05)

tmap_mode('plot')
```

Counting duplicates: jittering, adds a small perturbation to the duplicated points so they do not occupy the exact same space

```{r}
childcare_ppp_jit <- rjitter(childcare_ppp, 
                             retry=TRUE, 
                             nsim=1, 
                             drop=TRUE)
```

Check for any duplicated point in the geospatial data

```{r}
any(duplicated(childcare_ppp_jit))
```

## Creating owin object

When analysing spatial point patterns, it is a good practice to confine the analysis with a geographical area like Singapore boundary. In **spatstat**, an object called ***owin*** is specially designed to represent this polygonal region.

```{r}
sg_owin <- as(sg_sp, "owin")

plot(sg_owin)
```

```{r}
summary(sg_owin)
```

## Combining point events object and owin object

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]

summary(childcare_ppp)

plot(childcareSG_ppp)
```

## First-order Spatial Point Patterns Analysis

Kernel Density Estimation: the following code chunk computes a kernel density using the following configurations of density of spatstat

```{r}
kde_childcareSG_bw <- density(childcareSG_ppp,
                              sigma=bw.diggle,
                              edge=TRUE,
                            kernel="gaussian") 
```

Display the kernel density derived using plot()

```{r}
plot(kde_childcareSG_bw)
```

We realise that the density values of the output range is extremely small to comprehend. We will conduct rescaling of the KDE values in the subsequent step

Retrieve the bandwidth used to compute kde layer using code chunk

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

### Rescaling KDE

rescale() is used to convert the unit of measurement from meter to kilometer

```{r}
childcareSG_ppp.km <- rescale(childcareSG_ppp, 1000, "km")
```

then, we re-run density using the rescale data set and plot output kde map

```{r}
kde_childcareSG.bw <- density(childcareSG_ppp.km, sigma=bw.diggle, edge=TRUE, kernel="gaussian")
plot(kde_childcareSG.bw)
```

Difference from earlier version: data values on the legend

### Working with different automatic bandwidth methods

```{r}
 bw.CvL(childcareSG_ppp.km)
```

```{r}
bw.scott(childcareSG_ppp.km)
```

```{r}
bw.ppl(childcareSG_ppp.km)
```

```{r}
bw.diggle(childcareSG_ppp.km)
```

Comparing the output of using bw.diggle and bw.ppl methods

```{r}
kde_childcareSG.ppl <- density(childcareSG_ppp.km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
par(mfrow=c(1,2))
plot(kde_childcareSG.bw, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")
```

### Working with different kernel methods

By default, the kernel method used in *density.ppp()* is *gaussian*. But there are three other options, namely: Epanechnikov, Quartic and Dics. Computing 3 more kernel density estimations by using these 3 kernel functions

```{r}
par(mfrow=c(2,2))
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")
```

## Fixed and Adaptive KDE

Compute KDE layer by defining a bandwidth of 600 meter. Adaptive would be used when there are distinct clusters.

```{r}
kde_childcareSG_600 <- density(childcareSG_ppp.km, sigma=0.6, edge=TRUE, kernel="gaussian")
plot(kde_childcareSG_600)

#smoother as compared to kernel 
```

Computing KDE by using adaptive bandwidth

```{r}
kde_childcareSG_adaptive <- adaptive.density(childcareSG_ppp.km, method="kernel")
plot(kde_childcareSG_adaptive)
```

Comparing adaptive and fixed kernel density estimation outputs

```{r}
par(mfrow=c(1,2))
plot(kde_childcareSG.bw, main = "Fixed bandwidth")
plot(kde_childcareSG_adaptive, main = "Adaptive bandwidth")
```

Converting KDE output into grid object

```{r}
gridded_kde_childcareSG_bw <- as.SpatialGridDataFrame.im(kde_childcareSG.bw)
spplot(gridded_kde_childcareSG_bw)
```

Converting gridded output into raster

```{r}
kde_childcareSG_bw_raster <- raster(gridded_kde_childcareSG_bw)
```

Looking at properties of kde_childcareSG_bw_raster RasterLayer

```{r}
kde_childcareSG_bw_raster
```

### Assigning projection systems

```{r}
projection(kde_childcareSG_bw_raster) <- CRS("+init=EPSG:3414")
kde_childcareSG_bw_raster
```

## Visualising the output in tmap

```{r}
tm_shape(kde_childcareSG_bw_raster) + 
  tm_raster("v") +
  tm_layout(legend.position = c("right", "bottom"), frame = FALSE)
```

## Comparing spatial point patterns using KDE

```{r}
pg = mpsz[mpsz@data$PLN_AREA_N == "PUNGGOL",]
tm = mpsz[mpsz@data$PLN_AREA_N == "TAMPINES",]
ck = mpsz[mpsz@data$PLN_AREA_N == "CHOA CHU KANG",]
jw = mpsz[mpsz@data$PLN_AREA_N == "JURONG WEST",]
```

Plotting target planning areas

```{r}
par(mfrow=c(2,2))
plot(pg, main = "Ponggol")
plot(tm, main = "Tampines")
plot(ck, main = "Choa Chu Kang")
plot(jw, main = "Jurong West")
```

Converting the spatial point data frame into generic sp format

```{r}
pg_sp = as(pg, "SpatialPolygons")
tm_sp = as(tm, "SpatialPolygons")
ck_sp = as(ck, "SpatialPolygons")
jw_sp = as(jw, "SpatialPolygons")
```

Creating owin object

```{r}
pg_owin = as(pg_sp, "owin")
tm_owin = as(tm_sp, "owin")
ck_owin = as(ck_sp, "owin")
jw_owin = as(jw_sp, "owin")
```

Combining Childcare points and the study area

```{r}
childcare_pg_ppp = childcare_ppp_jit[pg_owin]
childcare_tm_ppp = childcare_ppp_jit[tm_owin]
childcare_ck_ppp = childcare_ppp_jit[ck_owin]
childcare_jw_ppp = childcare_ppp_jit[jw_owin]
```

Next, *rescale()* function is used to trasnform the unit of measurement from metre to kilometre.

```{r}
childcare_pg_ppp.km = rescale(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale(childcare_jw_ppp, 1000, "km")
```

The code chunk below is used to plot these four study areas and the locations of the childcare centres.

```{r}
par(mfrow=c(2,2))
plot(childcare_pg_ppp.km, main="Punggol")
plot(childcare_tm_ppp.km, main="Tampines")
plot(childcare_ck_ppp.km, main="Choa Chu Kang")
plot(childcare_jw_ppp.km, main="Jurong West")
```

Computing KDE

```{r}
par(mfrow=c(2,2))
plot(density(childcare_pg_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tempines")
plot(density(childcare_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="JUrong West")
```

Computing fixed bandwidth KDE

```{r}
par(mfrow=c(2,2))
plot(density(childcare_ck_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Chou Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="JUrong West")
plot(density(childcare_pg_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tampines")
```

## Nearest Neighbour Analysis

### Testing spatial point patterns using Clarks and Evans Test

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                nsim=99)
```

### Clark and Evans Test: Choa Chu Kang planning area

```{r}
clarkevans.test(childcare_ck_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

### Clark and Evans Test: Tampines planning area

```{r}
clarkevans.test(childcare_tm_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

## Second Order Spatial Point Process Using G-Function

### Choa Chu Kang Planning Area

Computing G-function estimation

```{r}
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim=c(0,500))
```

### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with G-fucntion

```{r}
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
```

```{r}
plot(G_CK.csr)
```

### Tampines planning area

Computing G-function estimation

```{r}
G_tm = Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Tampines are randomly distributed.

H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected is p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
```

```{r}
plot(G_tm.csr)
```

## Analysing Spatial Point Process Using F-funtion

### Choa Chu Kang planning area

```{r}
F_CK = Fest(childcare_ck_ppp)
plot(F_CK)
```

### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with F-fucntion

```{r}
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
```

```{r}
plot(F_CK.csr)
```

### Tampines planning area

```{r}
F_tm = Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

### Performing Complete Spatial Randomness Test

```{r}
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
```

```{r}
plot(F_tm.csr)
```

## Analysing Spatial Point Process Using K-Function

K-function measures the number of events found up to a given distance of any particular event.

### Choa Chu Kang Planning area

```{r}
K_ck = Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

### Tampines Planning area

```{r}
K_tm = Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . -r ~ r, 
     ylab= "K(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Tampines are randomly distributed.

H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_tm.csr, . - r ~ r, 
     xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

## Analysing Spatial Point Process Using L-Function

### Choa Chu Kang planning area

In graph: anything above line: clustering, below is regular

does not tell us any statistical analysis: have to use envelope method through simulations

```{r}
L_ck = Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value if smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```

if line is within simulation: fall within 95% confidence interval, although we observe clusters, it is not statistically significant. We fail to reject the null hypothesis. To reject null hypothesis, line must be above the envelope

### Tampines planning area

```{r}
L_tm = Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

### Performing complete spatial randomness test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Tampines are randomly distributed.

H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below will be used to perform the hypothesis testing.

```{r}
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_tm.csr, . - r ~ r, 
     xlab="d", ylab="L(d)-r", xlim=c(0,500))
```
