---
title: "In-class Excerise 4: Spatial Point Patterns Analysis"
date: "30 January 2023"
date-modified: "`r Sys.Date()`"
number-sections: true
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

## Getting Started

```{r}
pacman::p_load(maptools, sf, raster, spatstat, tmap)
```

Things to learn from this code chunk

```{r}
childcare_sf <- st_read("data/child-care-services-geojson.geojson") %>%
  st_transform(crs = 3414)
```

All spatial data: use projected coordinate system instead of geographic coordinate system.

```{r}
sg_sf <- st_read(dsn = "data", layer="CostalOutline")
```

Source is already in SVY21.

```{r}
mpsz_sf <- st_read(dsn = "data", 
                layer = "MP14_SUBZONE_WEB_PL")
```

## Visualising map

```{r}
tmap_mode("view")
          tm_shape(childcare_sf)+
            tm_dots(alpha=0.5, size=0.01)+
            tm_view(set.zoom.limits = c(11,14))# first number is the zoom out value, second number is the zoom in value ie, first number must always be smaller than the second number
          #we can use tm_bubbles if we are assigning specific values to each point
          
tmap_mode("plot")
```

## Data Wrangling for Spatstat

spatstat: needs conversion from sf to ppp. sf--\> spatial data frame--\> generic spatial object--\> ppp format.

```{r}
childcare <- as_Spatial(childcare_sf)
mpsz <- as_Spatial(mpsz_sf)
sg <- as_Spatial(sg_sf)
```

Converting spatial class into generic sp format

```{r}
childcare_sp <- as(childcare, "SpatialPoints")
sg_sp <- as(sg, "SpatialPolygons")
```

Convert generic sp format into ppp format

```{r}
childcare_ppp <- as(childcare_sp, "ppp")
childcare_ppp
```
